---
title: "GEO-TREES Tree Inventory Data Processing Workflow"
format: 
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{GEO-TREES Tree Inventory Data Processing Workflow}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    warning: false
    message: false
---

## Introduction

This purpose of this vignette is to demonstrate and document the complete
workflow for processing tree inventory data for the GEO-TREES initiative. The
code in this vignette matches the code in the R scripts elsewhere in this
repository. For automated processing of GEO-TREES tree inventory data it is
recommended to run the R scripts directly, not to use this vignette. Throughout
the vignette there are sign-posts to each R script.

Key steps:

* Format raw field data to match the GEO-TREES L0 data format
* Align taxonomic names
* Fit plot spatial polygons
* Estimate stem height, wood density, and biomass 
* Calculate summary statistics of forest structure and biomass stocks within 50x50 m quadrats 

This vignette uses an anonymised dataset, based on real data from a tropical
wet forest site in South America. To illustrate the use of all supported
auxiliary datasets and processing steps, some synthetic data have been
included. 

Synthetic datasets are provided for:

* Locally-measured wood density estimates

## Setup

Required packages:

```{r}
#| label: setup
#| warning: false
#| message: false
#| results: hide
library(dplyr)
library(tidyr)
library(readxl)
library(units)
library(sf)
library(BIOMASS)
library(brms)
library(yaml)
library(rocrateR)
```

## Create L0 data

This part of the workflow will vary among sites, depending on the structure of
the raw data. Consult the [GEO-TREES Tree Inventory Data Collection
Guidelines]() Section 16: _Data format_ for information on the GEO-TREES L0
data format.

We provide a number of functions to help create valid GEO-TREES L0 data products:

* `colCheck()` checks whether the columns in a dataframe exactly match those in a spec table based on their name and class (e.g. numeric, character).
* `valCheck()` runs a series of simple logical checks on the different GEO-TREES Tree Inventory L0 data objects. 
* `colGen()` creates empty columns in a dataframe if they don't already exist, using a spec table.
* `pasteVals()` intelligently pastes values together, replacing NAs with blanks, useful for combining values in the `code` column.

## Specify processing parameters

Choices made during the processing workflow, for example the choice to use a local or regional height-diameter estimation method, should be specified in a `param.yaml` YAML configuration file. This file also defines some site meta-data and filepaths. 

There should be one `param.yaml` file for each version of the dataset.

`param.yaml` files should contain the following parameters:

* `site_id` -- Name of the site, e.g. `"Bicuar"`
* `acquisition_id` -- ID of data acquisition, e.g. `"2024-08-03"`
* `out_dir` -- Path to where processed data will be exported, e.g. `"./dat/sites/Bicuar/PDA/2024-08-03/v1"`
* `raw_dir` -- Path to the raw input data, e.g. `"./dat/sites/Bicuar/PDA/2024-08-03/L0"`
* `s3_dir` -- Optional, path to AWS S3 directory containing raw input data, e.g. `"johngodlee/GEO-TREES_PDA/dat/sites/Amacayacu/L0"`
* `quad_dim` -- Dimensions in metres of quadrats, e.g. `[50, 50]`
* `opt_s3` -- Boolean (`true`/`false`). If true, raw input data will be pulled from AWS S3
* `product_version` -- String specifying the version of the data products, e.g. `"v1"`
* `height_method` -- Method used to estimate stem height, e.g. `"regional"`, `"local"`
* `wd_method` -- Method used to estimate wood density, e.g. `"regional"`, `"local"`

## Import L0 data

```{r}
# Define input data directory
indir <- "../extdata/L0"

# Define data output directory
outdir <- "./"

# Import stem measurements
s <- read.csv(file.path(indir, "stem.csv"))

# Import plot meta-data
p <- read.csv(file.path(indir, "plot.csv"))

# Import plot geo-location data
pt <- read.csv(file.path(indir, "plot_pt.csv"))

# Import taxonomy data (optional)
taxa <- read.csv(file.path(indir, "taxa.csv"))

# Import individual local wood density estimates (optional)
wd <- read.csv(file.path(indir, "wd.csv"))

# Import additional height-diameter measurements (optional)
hd <- read.csv(file.path(indir, "hd.csv"))

# TODO: Import local biomass allometry
# agb_allom <- readRDS(file.path(indir, "agb_allom.rds"))

# TODO: Import local height-diameter allometry
# hd_allom <- readRDS(file.path(indir, "hd_allom.rds"))
```

## Correct taxonomy 

`02_taxa.R`

```{r}
# Combine taxa across tables
taxon_vec <- sort(unique(c(
  stem$taxon_name,
  wd$taxon_name,
  height$taxon_name)))

# Check names
taxa <- BIOMASS::correctTaxo(
  genus = taxon_vec, 
  species = NULL,
  interactive = TRUE, 
  preferAccepted = TRUE,
  preferFuzzy = FALSE,
  sub_pattern = subPattern(),
  useCache = TRUE,
  useAPI = FALSE,
  capacity = 60,
  fill_time_s = 1, 
  timeout = 60)

# Write stem taxonomic information to file
write.csv(taxa, file.path(outdir, "stem_taxa.csv"), row.names = FALSE)

# Write WFO cache to file
saveRDS(BIOMASS:::the$wfo_cache, file.path(outdir, "wfo_cache.rds"))
```

## Fit plot spatial polygons

`03_quad.R`

```{r}
# Extract plot corners
plot_check <- BIOMASS::plotPolygonFit(
  point_data = plot_pt,
  method = "procrustes",
  rel_col = c("x_rel_m", "y_rel_m"),
  proj_col = c("rover_easting_utm_m", "rover_northing_utm_m"),
  lonlat_col = NULL,
  corner_col = "corner",
  plot_col = "plot_id",
  max_dist = 10,
  rm_outliers = TRUE,
  tree_data = stem[stem$plot_id %in% plot_pt$plot_id,],
  tree_rel_col = c("x_rel_m", "y_rel_m"),
  tree_plot_col = "plot_id",
  raster = NULL,
  shapefile = NULL,
  prop_tree = NULL,
  threshold_tree = NULL, 
  draw_plot = FALSE,
  ask = FALSE)

# Divide plot
plot_divide <- BIOMASS::divide_plot(
  corner_data = plot_check$corner_coord,
  rel_coord = c("x_rel", "y_rel"),
  proj_coord = c("x_proj", "y_proj"),
  longlat = NULL,
  grid_size = param$quad_dim,
  grid_tol = 1,
  origin = NULL,
  tree_data = plot_check$tree_data,
  tree_coords = c("x_rel", "y_rel"),
  corner_plot_ID = "plot_ID",
  tree_plot_ID = "plot_ID",
  sd_coord = NULL, 
  n = 100)

# Create plot polygons 
plot_poly <- plot_check$polygon %>% 
  rename(plot_id = plot_ID) %>% 
  st_set_crs(unique(plot_pt$crs_epsg)) %>% 
  mutate(    
    site_id = param$site_id,
    acquisition_id = param$acquisition_id, 
    .before = everything())

# Create quadrat polygons
quad_pt <- plot_divide$sub_corner_coord %>% 
  st_as_sf(., coords = c("x_proj", "y_proj"), crs = unique(plot_pt$crs_epsg)) %>% 
  mutate(    
    site_id = param$site_id,
    acquisition_id = param$acquisition_id, 
    .before = everything()) %>% 
  rename(
    plot_id = plot_ID, 
    quadrat_id = subplot_ID,
    x_rel_m = x_rel,
    y_rel_m = y_rel)

quad_poly <- quad_pt %>% 
  group_by(site_id, acquisition_id, plot_id, quadrat_id) %>%
  summarise() %>% 
  st_convex_hull() %>%
  ungroup()

# Create stem points
stem_pt <- plot_divide$tree_data %>% 
  filter(!is.na(x_proj), !is.na(y_proj)) %>% 
  st_as_sf(., coords = c("x_proj", "y_proj"), crs = unique(plot_pt$crs_epsg)) %>% 
  dplyr::select(
    record_id,
    quadrat_id = subplot_ID)

# Write plot polygons to file
st_write(plot_poly, file.path(outdir, "plot_poly.gpkg"), delete_dsn = TRUE)

# Write plot corner points
st_write(plot_pt, file.path(outdir, "plot_corner.gpkg"), delete_dsn = TRUE)

# Write quadrat polygons to file
st_write(quad_poly, file.path(outdir, "quad_poly.gpkg"), delete_dsn = TRUE)

# Write quadrat points to file
st_write(quad_pt, file.path(outdir, "quad_corner.gpkg"), delete_dsn = TRUE)

# Write global stem coordinates to file
st_write(stem_pt, file.path(outdir, "stem_pt.gpkg"), delete_dsn = TRUE)

# Write output of divide_plot() to file
saveRDS(plot_divide, file.path(outdir, "plot_divide.rds"))
```

## Estimate stem wood density

`04_wd.R`

```{r}
# Join taxonomy data to stem data
stem_taxa_all <- left_join(stem, taxa, by = c("taxon_name" = "nameOriginal"))

# Optionally process local wood density data
if (param$wd_method == "field") { 
  wd_summ <- wd %>% 
    left_join(., taxa, by = c("taxon_name" = "nameOriginal")) %>% 
  group_by(
    family = familyAccepted,
    genus = genusAccepted,
    species = speciesAccepted) %>% 
  summarise(
    meanWD = mean(wd_gcm3, na.rm = TRUE),
    sdWD = sd(wd_gcm3, na.rm = TRUE))
}

# Estimate wood density for each stem measurement
wd_all <- BIOMASS::getWoodDensity(
  genus = stem_taxa_all$genusAccepted, 
  species = stem_taxa_all$speciesAccepted,
  stand = stem_taxa_all$plot_id,
  family = stem_taxa_all$familyAccepted,
  addWoodDensity = if (param$wd_method == "field") { wd_summ },
  verbose = TRUE)

# Create output dataframe
wd_out <- cbind(record_id = stem_taxa_all$record_id, wd_all)

# Write to file
write.csv(wd_out, file.path(outdir, "stem_wd.csv"), row.names = FALSE)
```

## Estimate stem height

`05_height.R`

Regional height estimation where no height data are available

Local height estimation where height data are available

Local height estimation where existing height-diameter allometry is available

## Estimate stem AGB 

`06_agb_stem.R`

```{r}
# Estimate stem-level AGB
stem_agb <- stem %>% 
  left_join(., stem_wd, by = "record_id") %>% 
  left_join(., stem_height, by = "record_id") %>% 
  mutate(
    agb_Mg = BIOMASS::computeAGB(
      allometry = chave2014,
      D = .$diam_cm,
      WD = .$meanWD,
      H = .$height_m_pred)[[1]],
    ba_m2 = base::pi * (.$diam_cm / 2)^2 / 10000,
    volume_m3 = agb_Mg / meanWD) %>% 
  dplyr::select(
    record_id, 
    agb_Mg,
    ba_m2,
    volume_m3)

# Write stem-level AGB to file
write.csv(stem_agb, file.path(outdir, "stem_agb.csv"), row.names = FALSE)
```

## Filter stem data 

`07_record_fil.R`

```{r}
# Filter stem data
record_fil <- stem %>% 
  left_join(., stem_pt, by = "record_id") %>% 
  left_join(., plot[,c("plot_id", "meas_diam_min_cm")], 
    by = "plot_id") %>% 
  filter(
    !is.na(diam_cm),
    !is.na(quadrat_id),
    diam_cm >= meas_diam_min_cm,
    grepl("A", code),
    grepl("S", code),
    !grepl("B", code),
    !grepl("T", code),
    !grepl("M", code)
  ) %>% 
  group_by(site_id, plot_id, tree_id, stem_id) %>% 
  slice_max(
    order_by = tibble(pom_m, measurement_date, diam_cm), 
    n = 1, 
    with_ties = FALSE) %>%
  ungroup() %>% 
  pull(record_id) %>% 
  as.character()

# Write filtered stem data to file
writeLines(record_fil, file.path(outdir, "record_fil.txt"))
```

## Estimate quadrat AGBD and propagate error

`08_agb_mc.R`

```{r}
# Extract plot centres
p_cent <- plot_poly %>% 
  st_transform(., crs = 4326) %>% 
  group_by(site_id, plot_id, acquisition_id) %>% 
  summarise() %>% 
  st_centroid() %>% 
  cbind(., st_coordinates(.)) %>% 
  st_drop_geometry() %>% 
  dplyr::select(plot_id, X, Y)

# Filter stems data
stem_fil <- stem %>% 
  left_join(., p_cent, by = "plot_id") %>% 
  left_join(., stem_wd, by = "record_id") %>% 
  left_join(., stem_pt, by = "record_id") %>% 
  filter(record_id %in% record_fil)

# Split by quadrat
stem_split <- split(stem_fil, stem_fil$quadrat_id)

# Define number of simulations
nsim <- 1000

# Run AGB MC error propagation
quad_length <- length(stem_split)
quad_agb_mc_list <- lapply(seq_along(stem_split), function(x) { 
  message(paste0(x, " / ", quad_length, " - ", names(stem_split)[x]))
  if (nrow(stem_split[[x]]) < 2) {
    out <- list(
      "meanAGB" = stem_split[[x]]$agb_Mg,
      "medAGB" = stem_split[[x]]$agb_Mg,
      "sdAGB" = NA_real_,
      "credibilityAGB" = c("2.5%" = NA_real_, "97.5%" = NA_real_),
      "AGB_simu" = NA_real_)
  } else {
    out <- BIOMASS::AGBmonteCarlo(
      D = stem_split[[x]]$diam_cm,
      WD = stem_split[[x]]$meanWD,
      coord = st_drop_geometry(stem_split[[x]][,c("X", "Y")]),
      Dpropag = "chave2004",
      errWD = stem_split[[x]]$sdWD,
      n = nsim)
    # here add HDmodel (from 05 step) so that all uncertainties are propagated
  }
})
names(quad_agb_mc_list) <- names(stem_split)

# Sum stem-level AGB simulations, to get AGBD per quadrat per simulation
quad_agb_simu <- bind_rows(lapply(names(quad_agb_mc_list), function(x) { 
  out <- data.frame(
    quadrat_id = x,
    sim = seq_len(nsim),
    sumAGB = colSums(as.matrix(quad_agb_mc_list[[x]]$AGB_simu))
  )
  rownames(out) <- NULL
  out
}))
# here can be done with BIOMASS::subplot_summary and thus includes uncertainties 
# on coordinates, needs divide_plot outputs from 03_quad

# Write quadrat simulations to file
write.csv(quad_agb_simu, file.path(outdir, "quad_agb_mc.csv"), row.names = FALSE)

# Calculate mean and standard deviation of stem AGB
stem_agb_mc <- bind_rows(lapply(seq_along(quad_agb_mc_list), function(x) {
  simu <- quad_agb_mc_list[[x]]$AGB_simu
  if (is.matrix(simu)) {
    agb_Mg_mean <- apply(simu, 1, mean)
    agb_Mg_sd <- apply(simu, 1, sd)
  } else {
    agb_Mg_mean <- NA_real_
    agb_Mg_sd <- NA_real_
  }
  data.frame(record_id = stem_split[[x]]$record_id, agb_Mg_mean, agb_Mg_sd)
}))

# Write stem-level data to file
write.csv(stem_agb_mc, file.path(outdir, "stem_agb_mc.csv"), row.names = FALSE)

# Extract summary statistics from AGB MC error propagation simulations
quad_agb_mc_summ <- bind_rows(lapply(names(quad_agb_mc_list), function(x) { 
  data.frame(
    quadrat_id = x,
    agb_Mg_sum_mc_mean = quad_agb_mc_list[[x]]$meanAGB,
    agb_Mg_sum_mc_median = quad_agb_mc_list[[x]]$medAGB,
    agb_Mg_sum_mc_sd = quad_agb_mc_list[[x]]$sdAGB,
    agb_Mg_sum_mc_se = quad_agb_mc_list[[x]]$sdAGB / nsim,
    agb_Mg_sum_mc_ci2.5 = unname(quad_agb_mc_list[[x]]$credibilityAGB[1]),
    agb_Mg_sum_mc_ci97.5 = unname(quad_agb_mc_list[[x]]$credibilityAGB[2]))
}))

# Write quadrat-level summary data to file
write.csv(quad_agb_mc_summ, file.path(outdir, "quad_agb.csv"), row.names = FALSE)
```

## Compile GEO-TREES L1, L2, and L3 data products 

`09_stem_summ.R`
`10_quad_summ.R`
`11_brm.R`

Create ROCrate files.

* L1 re-formatted stem measurements.
* L1 re-formatted plot geo-location points
* L1 re-formatted plot meta-data
* L2 stem AGB estimates.
* L1 plot polygons.
* L3 AGBD estimates within plot quadrats.
